% vmdbench_workshop.tex — short workshop / position paper draft.
% Compiles with plain pdflatex (only standard packages). To submit, drop the
% body into the venue's style file (NeurIPS/ICML/etc.) and move authors/abstract.
\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{times}
\usepackage{booktabs}
\usepackage{microtype}
\usepackage{amsmath}
\usepackage[hidelinks]{hyperref}
\usepackage{xcolor}
\usepackage{titlesec}
\titlespacing*{\section}{0pt}{1.1em}{0.4em}
\titlespacing*{\subsection}{0pt}{0.8em}{0.3em}
\setlength{\parskip}{0.25em}

\title{\vspace{-3em}Completion Is Not Correctness:\\
Verifying LLM Agents for Molecular Visualization}
\author{Peter [Surname]\thanks{Corresponding author.} \quad Diego [Surname] \\
\small Theoretical and Computational Biophysics Group / Beckman Institute \\
\small \texttt{pinhao.peter@gmail.com}}
\date{\today}

\begin{document}
\maketitle

\begin{abstract}
\noindent
Large language model (LLM) agents can now drive scientific visualization tools such as
VMD by emitting Tcl commands, and recent benchmarks score them by \emph{task completion}:
did the agent produce an answer or render a scene? We argue that for scientific use this is
the wrong target. We present \textbf{vmdbench}, a correctness-oriented benchmark that checks
each agent answer against an independently computed \emph{oracle} value (cross-validated
against MDAnalysis and Biopython), and we use it to measure the gap between completion and
correctness for an open-weight agent (Qwen2.5-72B). The agent almost always \emph{completes}
tasks while being correct far less often --- it confidently reports plausible-but-wrong numbers
and builds empty visualizations that pass a hollow rubric. We then show that three targeted
interventions --- proactive reference injection, Tcl pre-validation, and semantic measurement
tools --- raise correctness to 97\% and collapse the completion--correctness gap to roughly
zero on the analysis suite. Finally, through a held-out evaluation on operations the tools do
\emph{not} cover, we decompose that gain honestly: most of it is tool coverage rather than a
general capability jump, with a smaller, idiom-specific improvement attributable to injection.
We release the benchmark, oracles, and agent harness.
\end{abstract}

\section{Introduction}
LLM agents are increasingly proposed as natural-language front-ends to scientific software:
a researcher asks for a radius of gyration or a licorice representation, and the agent writes
the Tcl or Python that the tool executes. Evaluations of such agents, including
SciVisAgentBench-style suites, predominantly reward \emph{completion} --- the agent ran to
termination and produced \emph{an} answer or \emph{a} rendered image. For a chatbot this is a
reasonable proxy. For a scientific instrument it is dangerous: a confidently reported wrong
number is worse than a refusal, and a visualization that ``looks done'' but contains zero
representations is a silent failure.

This paper makes a simple, measurable point: \textbf{completion is not correctness}, and the
distance between them is large enough to change conclusions. We build a benchmark that measures
correctness directly, quantify the gap for a strong open-weight agent, and then close most of it
--- while being explicit about \emph{why} it closes, separating genuine generalization from the
mechanical effect of giving the agent tools that happen to match the benchmark's metrics.

Our contributions are: (i) \textbf{vmdbench}, a correctness benchmark for VMD agents built on
independently computed, cross-validated oracle values for both analysis and visualization tasks;
(ii) a measurement of the \emph{completion--correctness gap} for Qwen2.5-72B across five protein
structures; (iii) three interventions that raise correctness to 97\%; and (iv) a held-out rigor
study that decomposes the improvement into tool coverage versus generalization.

\section{vmdbench: measuring correctness, not completion}
Each task pairs a natural-language prompt with a deterministic \emph{oracle}: a script that
computes the gold value directly from the structure, independent of the agent. For analysis
tasks (radius of gyration, SASA, atom and residue counts, $C\alpha$ distances, RMSD) the oracle
emits a scalar that we compare to the agent's answer within a metric-specific tolerance. To
guard against oracle bugs we cross-validate the gold against two independent libraries,
MDAnalysis and Biopython; 19 of 20 metrics agree to tolerance, and the lone discrepancy is a
known definitional difference rather than an error.

Crucially, we also verify \emph{visualization} tasks, where completion-style scoring is weakest.
Instead of asking ``did an image appear,'' we assert on the resulting scene state --- whether the
requested representation actually exists, whether the named style and coloring were applied, and
whether a non-empty image was written. This catches the failure mode where an agent claims to
have built a ``licorice representation'' but created zero representations.

\section{The completion--correctness gap}
We evaluate Qwen2.5-72B (AWQ 4-bit, served locally) as a VMD agent across five structures
ranging from 327 to over 9{,}000 atoms. The agent is a near-perfect \emph{completer}: it
terminates and reports a value on almost every task. Its \emph{correctness}, however, is far
lower. The agent recomputes self-RMSD as a nonzero value when it should be exactly zero, reports
plausible but out-of-tolerance radii of gyration, and --- on visualization tasks --- produces
scenes that satisfy a hollow rubric while containing no representations. The gap between
completion and correctness is not noise; it is the central measurement, and a benchmark that
scores only completion would report this agent as strong.

\section{Closing the gap}
Three interventions, each motivated by an observed failure mode, move the agent from
``completes everything, correct sometimes'' to ``correct when it answers.''

\paragraph{(1) Proactive reference injection.} The agent rarely retrieves documentation even
when instructed to. Rather than rely on agent-initiated retrieval, we inject a small, curated
VMD-Tcl cheat-sheet into the context up front (a DocPrompting-style move). This fixes idiom-level
errors such as the self-RMSD recomputation.

\paragraph{(2) Tcl pre-validation.} The agent occasionally emits syntactically incomplete Tcl
(unbalanced braces or brackets), which can hang the interpreter. We reject such commands before
execution with a lightweight balance check and raise per-command timeouts.

\paragraph{(3) Semantic measurement tools.} We expose typed tools (\texttt{vmd\_measure},
\texttt{vmd\_represent}) that emit the correct Tcl for a named operation, so the model only has to
choose the operation, not author the script.

Together these raise correctness to \textbf{97\%} on the analysis suite, with the
completion--correctness gap collapsing to approximately zero: when the improved agent answers, it
is right.

\section{What actually generalizes}
A 97\% number invites an obvious objection: the semantic tools' operations \emph{are} the
benchmark's metrics, so some of the gain is the tool doing the task by construction. We test this
directly on a \textbf{held-out} set of six operations the tools do \emph{not} cover --- hydrogen
bonds, a three-atom angle, a backbone dihedral, center of mass, bounding-box extent, and a
filtered glycine count --- requiring the agent to write raw Tcl. We compare three arms: the raw
agent (\textsc{none}), reference injection only (\textsc{inject}), and the full system
(\textsc{+tools}); each cell is a pass rate over two structures and three seeds (Table~\ref{tab:heldout}).

\begin{table}[h]
\centering
\small
\begin{tabular}{lccc}
\toprule
held-out operation & \textsc{none} & \textsc{inject} & \textsc{+tools} \\
\midrule
hydrogen bonds        & 6/6 & 6/6 & 6/6 \\
3-atom angle          & 4/6 & 6/6 & 6/6 \\
dihedral ($\phi$)     & 3/6 & 6/6 & 6/6 \\
center of mass        & 6/6 & 6/6 & 6/6 \\
bounding-box extent   & 2/6 & 0/6 & 0/6 \\
glycine count         & 0/6 & 0/6 & 0/6 \\
\midrule
\textbf{total}        & \textbf{21/36} & \textbf{24/36} & \textbf{24/36} \\
\bottomrule
\end{tabular}
\caption{Held-out correctness on tool-uncovered operations. \textsc{inject} equals \textsc{+tools},
so the semantic tools add nothing off their coverage; the only generalization is from injection,
and only on the idioms the cheat-sheet teaches.}
\label{tab:heldout}
\end{table}

Three things follow. First, the headline 97\% is \emph{largely tool coverage}: stripped of the
covering tools, correctness falls to 67\% (24/36). Second, the semantic tools add \emph{nothing}
off their coverage --- \textsc{inject} and \textsc{+tools} are identical --- so they are an
engineering win when they cover the operation and neutral otherwise. Third, the real
generalization comes from \emph{injection} (21$\to$24), and only on the idioms the cheat-sheet
teaches: the three-atom angle (4$\to$6) and the dihedral (3$\to$6). Operations the cheat-sheet
never mentions are either solved regardless (hydrogen bonds, center of mass) or resist every arm
(glycine count). Injection is not free: it slightly \emph{hurt} the bounding-box task (2$\to$0),
evidence that an injected reference can distract on operations it does not cover.

\section{Limitations}
The held-out study uses two structures and three seeds per cell; the small denominators mean
single-task movements (e.g., bounding-box extent) should be read as suggestive, not definitive.
Our oracles encode specific operational definitions (e.g., hydrogen-bond geometry, RMSD against
self), and an agent using a different-but-valid convention may be scored wrong; we mitigate this
with cross-validation but do not eliminate it. Finally, results are reported for a single
open-weight model; multi-model baselines are in progress and would strengthen the claim that the
completion--correctness gap is a property of the task formulation rather than of one model.

\section{Conclusion}
For scientific agents, scoring completion rewards confident wrongness. We show the gap is large
and measurable, that targeted interventions can close it on covered operations, and --- being
honest about it --- that most of the closing is tool coverage rather than a general capability
jump. We advocate correctness-oriented, oracle-backed evaluation as the default for agents that
drive scientific instruments, and release vmdbench toward that end.

\paragraph{Reproducibility.} The benchmark, oracle scripts, cross-validation harness, agent
adapter, and held-out runner are released with deterministic seeds and per-task gold values.

\end{document}
